---
title: "Data Viz Tips & Tricks"
author: "Eric Ekholm"
format: revealjs
---

# About Me

- Data Specialist at CCPS
    - formerly VDOE & MS English Teacher
- PhD in Educational Psychology
- 2 kids, 1 pitbull
- Enjoy data, coding, reading, and crossfit
- [Twitter](https://twitter.com/ekholm_e), [Github](https://github.com/ekholme), [Website](https://www.ericekholm.com)

# About These Materials

- All plots created using [R](https://www.r-project.org/) and the [`ggplot2`](https://ggplot2.tidyverse.org/) visualization library
- Slides created using [Quarto](https://quarto.org/)
- Data is publicly available from VDOE (& is also available in this repository).
- All code/slides available in this [Github repository](https://github.com/ekholme/data_viz_tips)

# Use Data to Tell a Story

- Embrace the fact that visualizations are subjective
- Make your viz decisions shape the story you want to tell
- Tailor the story you're telling to the decisions you're trying to inform

# Tip #1: Tell Viewers the Takeaway

```{r}
#| echo: false

library(tidyverse)
library(ccpsr)
library(janitor)

theme_set(theme_ore())

#set color hexes
herm <- "#660012"
g80 <- "#cccccc"

#read in and lightly manipulate some data
sol_raw <- read_csv(here::here("data/sol_longitudinal_results.csv")) |>
    clean_names() |>
    mutate(school_year = as.numeric(str_replace_all(school_year, "(^\\d{4}).*", "\\1")))

fm_df <- read_csv(here::here("data/fall_membership_statistics.csv")) |>
    clean_names()

top_nms <- fm_df |>
    slice_max(order_by = total_count, n = 10) |>
    pull(division_number)

sol_top  <- sol_raw |>
    filter(division_number %in% top_nms)

ccps_rd3 <- sol_raw |>
    filter(test == "English Reading", test_level == "Grade 3", division_number == 21)


```

# Generic Title

```{r}
basep <- ccps_rd3 |>
    ggplot(aes(x = school_year, y = pass_rate)) +
    geom_line(size = 2) +
    theme(
        panel.grid.minor = element_blank()
    )

basep +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        caption = "Data: VDOE Test Results Build-a-Table",
        title = "CCPS 3rd Grade Reading SOL Pass Rates Over Time"
    )
```

# Opinionated Title

```{r}
basep +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        caption = "Data: VDOE Test Results Build-a-Table",
        title = "CCPS 3rd Grade Reading Scores Decline Steeply after 2015"
    )
```

# Tip #2: Avoid Extraneous Information

# Too Much

```{r}
 rd3_top <- sol_top |>
    filter(test == "English Reading", test_level == "Grade 3")

 rd3_top |>
    ggplot(aes(x = school_year, y = pass_rate, color = division_name)) +
    geom_line() +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        title = "Grade 3 Reading SOL Pass Rates for VA's 10 Largest Divisions",
        caption = "Data: VDOE Test Results Build-a-Table"
    ) +
    scale_color_discrete(name = "Division")
```

# Just Enough

```{r}
 rd3_top |>
    filter(division_number != 21) |>
    ggplot() +
    geom_line(aes(x = school_year, y = pass_rate, group = division_number), color = g80, size = 1) +
    geom_line(data = ccps_rd3, aes(x = school_year, y = pass_rate), size = 2, color = herm) +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        title = "<span style='color:#660012'>CCPS's</span> 3rd Grade Reading Trends Largely Mirrored those of Other Large Divisions<br>in VA",
        caption = "Data: VDOE Test Results Build-a-Table"
    )

```

# Tip #3: Build to Complex Visuals

- Generally, try to keep visualizations simple and aligned with your key takeaways
- If you need to do something more complex, build to it over several plots

# Start With

```{r}
ccps_rd3 |>
    ggplot(aes(x = school_year, y = pass_rate)) +
    geom_line(size = 2, color = herm) +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        caption = "Data: VDOE Test Results Build-a-Table",
        title = "<span style='color:#660012'>CCPS's</span> 3rd Grade Reading Scores Decline Steeply after 2015"
    )
```

# Then

```{r}
 rd3_top |>
    filter(division_number != 21) |>
    ggplot() +
    geom_line(aes(x = school_year, y = pass_rate, group = division_number), color = g80, size = 1) +
    geom_line(data = ccps_rd3, aes(x = school_year, y = pass_rate), size = 2, color = herm) +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        title = "<span style='color:#660012'>CCPS's</span> 3rd Grade Reading Trends Largely Mirrored those of Other Large Divisions<br>in VA",
        caption = "Data: VDOE Test Results Build-a-Table"
    )
```

# Finally

```{r}
gs <- paste0("Grade ", 3:8)

rd_es_ms_top <- sol_top |>
    filter(test == "English Reading", test_level %in% gs) |>
    mutate(is_ccps = if_else(division_number == 21, herm, g80),
            division_name = if_else(str_detect(division_name, "^Chester"), str_replace_all(division_name, "^", "z"), division_name)) |>
    arrange(division_name)

ccps_es_ms_rd <- rd_es_ms_top |>
    filter(division_number == 21)

rd_es_ms_top |>
    #filter(division_number != 21) |>
    ggplot() +
    geom_line(aes(x = school_year, y = pass_rate, group = division_name, color = is_ccps), size = 1) +
    # geom_line(data = ccps_rd3, aes(x = school_year, y = pass_rate), size = 2, color = herm) +
    facet_wrap(vars(test_level)) +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        title = "<span style='color:#660012'>CCPS's</span> Grade 3-8 Reading Trends Largely Mirrored those of Other Large Divisions<br>in VA",
        caption = "Data: VDOE Test Results Build-a-Table"
    ) +
    scale_color_identity()
```