---
title: "Data Viz Tips & Tricks"
author: "Eric Ekholm"
format: revealjs
---

# About Me

- Data Specialist at CCPS
    - formerly VDOE & MS English Teacher
- PhD in Educational Psychology
- 2 kids, 1 pitbull
- Enjoy data, coding, reading, and crossfit
- [Twitter](https://twitter.com/ekholm_e), [Github](https://github.com/ekholme), [Website](https://www.ericekholm.com)

# About These Materials

- All plots created using [R](https://www.r-project.org/) and the [`ggplot2`](https://ggplot2.tidyverse.org/) visualization library
- Slides created using [Quarto](https://quarto.org/)
- Data is publicly available from VDOE (& is also available in this repository).
- All code/slides available in this [Github repository](https://github.com/ekholme/data_viz_tips)

# Use Data to Tell a Story

- Embrace the fact that visualizations are subjective
- Make your viz decisions shape the story you want to tell
- Tailor the story you're telling to the decisions you're trying to inform

# Tip #1: Tell Viewers the Takeaway

```{r}
#| echo: false

library(tidyverse)
library(ccpsr)
library(janitor)

theme_set(theme_ore())

sol_raw <- read_csv(here::here("data/sol_longitudinal_results.csv")) |>
    clean_names() |>
    mutate(school_year = as.numeric(str_replace_all(school_year, "(^\\d{4}).*", "\\1")))

ccps_rd3 <- sol_raw |>
    filter(test == "English Reading", test_level == "Grade 3", division_number == 21)
```

# Generic Title

```{r}
basep <- ccps_rd3 |>
    ggplot(aes(x = school_year, y = pass_rate)) +
    geom_line(size = 2) +
    theme(
        panel.grid.minor = element_blank()
    )

basep +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        caption = "Data: VDOE Test Results Build-a-Table",
        title = "CCPS 3rd Grade Reading SOL Pass Rates Over Time"
    )
```

# Opinionated Title

```{r}
basep +
    labs(
        x = "School Year (Fall)",
        y = "Pass Rate (%)",
        caption = "Data: VDOE Test Results Build-a-Table",
        title = "CCPS 3rd Grade Reading Scores Decline Steeply after 2015"
    )
```

